RProtoBuf: Efficient Cross-Language Data Serialization in R

### 2013-12-17 papers/rjournal/eddelbuettel-francois-stokely.Rnw
Move over some sections from the introduction vignette that can be included here.  I've started moving the pieces over piecemeal because the intro vignette is too much like a reference manual and needs more of a narrative which I'm attempting to provide here.
Specifically, section 3 "Classes, Methods, and Pseudo-Methods" of the
intro vignette is 20 pages on its own of reference material that would
not be as appropriate for a writeup in R Journal or JSS or something.
It is good content, but we need to replace much of that with very
concise tables or similar.

The number of classes and methods i

@@

% Example code snippet.
% TODO(mstokely): Remove this.
\begin{example}
  x <- 1:10
  result <- myFunction(x)
\end{example}

@@

@@

@@

This file is only a basic article template. For full details of \emph{The R Journal} style and information on how to prepare your article for submission, see the \href{http://journal.r-project.org/latex/RJauthorguide.pdf}{Instructions for Authors}.

>>>>

\emph{TODO(mstokely): Remove this example code snippet}

\begin{example}
  x <- 1:10
  result <- myFunction(x)
\end{example}

@@

<<>>=
library(RProtoBuf)
args( readProtoFiles )
@

Using the \texttt{file} argument, one can specify one or several file
paths that ought to be proto files.

<<>>=
proto.dir <- system.file( "proto", package = "RProtoBuf" )
proto.file <- file.path( proto.dir, "addressbook.proto" )
<<eval=FALSE>>=
readProtoFiles( proto.file )
@

With the \texttt{dir} argument, which is
ignored if the \texttt{file} is supplied, all files matching the
\texttt{.proto} extension will be imported.

<<>>=
dir( proto.dir, pattern = "\\.proto$", full.names = TRUE )
<<eval=FALSE>>=
readProtoFiles( dir = proto.dir )
@

Finally, with the
\texttt{package} argument (ignored if \texttt{file} or
\texttt{dir} is supplied), the function will import all \texttt{.proto}
files that are located in the \texttt{proto} sub-directory of the given
package. A typical use for this argument is in the \texttt{.onLoad}
function of a package.

<<eval=FALSE>>=
readProtoFiles( package = "RProtoBuf" )
@

Once the proto files are imported, all message descriptors are
are available in the R search path in the \texttt{RProtoBuf:DescriptorPool}
special environment. The underlying mechanism used here is
described in more detail in section~\ref{sec-lookup}.

<<>>=
ls( "RProtoBuf:DescriptorPool" )
@

\subsection{Creating a message}

The objects contained in the special environment are
descriptors for their associated message types. Descriptors will be
discussed in detail in another part of this document, but for the
purpose of this section, descriptors are just used with the \texttt{new}
function to create messages.

<<>>=
p <- new( tutorial.Person, name = "Romain", id = 1 )
@

\subsection{Access and modify fields of a message}

Once the message is created, its fields can be queried
and modified using the dollar operator of R, making protocol
buffer messages seem like lists.

<<>>=
p$name
p$id
p$email <- "francoisromain@free.fr"
@

However, as opposed to R lists, no partial matching is performed
and the name must be given entirely.

The \verb|[[| operator can also be used to query and set fields
of a mesages, supplying either their name or their tag number :

<<>>=
p[["name"]] <- "Romain Francois"
p[[ 2 ]] <- 3
p[[ "email" ]]
@

Protocol buffers include a 64-bit integer type, but R lacks native
64-bit integer support.  A workaround is available and described in
Section~\ref{sec:int64} for working with large integer values.

% TODO(mstokely): Document extensions here.
% There are none in addressbook.proto though.

\subsection{Display messages}

Protocol buffer messages and descriptors implement \texttt{show}
methods that provide basic information about the message :

<<>>=
p
@

For additional information, such as for debugging purposes,
the \texttt{as.character} method provides a more complete ASCII
representation of the contents of a message.

<<>>=
writeLines( as.character( p ) )
@

\subsection{Serializing messages}

However, the main focus of protocol buffer messages is
efficiency. Therefore, messages are transported as a sequence
of bytes. The \texttt{serialize} method is implemented for
protocol buffer messages to serialize a message into the sequence of
bytes (raw vector in R speech) that represents the message.

<<>>=
serialize( p, NULL )
@

The same method can also be used to serialize messages to files :

<<>>=
tf1 <- tempfile()
tf1
serialize( p, tf1 )
readBin( tf1, raw(0), 500 )
@

Or to arbitrary binary connections:

<<>>=
tf2 <- tempfile()
con <- file( tf2, open = "wb" )
serialize( p, con )
close( con )
readBin( tf2, raw(0), 500 )
@

\texttt{serialize} can also be used in a more traditionnal
object oriented fashion using the dollar operator :

<<>>=
# serialize to a file
p$serialize( tf1 )
# serialize to a binary connection
con <- file( tf2, open = "wb" )
p$serialize( con )
close( con )
@

\subsection{Parsing messages}

The \texttt{RProtoBuf} package defines the \texttt{read}
function to read messages from files, raw vector (the message payload)
and arbitrary binary connections.

<<>>=
args( read )
@

The binary representation of the message (often called the payload)
does not contain information that can be used to dynamically
infer the message type, so we have to provide this information
to the \texttt{read} function in the form of a descriptor :

<<>>=
message <- read( tutorial.Person, tf1 )
writeLines( as.character( message ) )
@

The \texttt{input} argument of \texttt{read} can also be a binary
readable R connection, such as a binary file connection:

<<>>=
con <- file( tf2, open = "rb" )
message <- read( tutorial.Person, con )
close( con )
writeLines( as.character( message ) )
@

Finally, the payload of the message can be used :

<<>>=
# reading the raw vector payload of the message
payload <- readBin( tf1, raw(0), 5000 )
message <- read( tutorial.Person, payload )
@

\texttt{read} can also be used as a pseudo method of the descriptor
object :

<<>>=
# reading from a file
message <- tutorial.Person$read( tf1 )
# reading from a binary connection
con <- file( tf2, open = "rb" )
message <- tutorial.Person$read( con )
close( con )
# read from the payload
message <- tutorial.Person$read( payload )
@

@@

\section{Descriptor lookup}
\label{sec-lookup}

The \texttt{RProtoBuf} package uses the user defined tables framework
that is defined as part of the \texttt{RObjectTables} package available
from the OmegaHat project.

The feature allows \texttt{RProtoBuf} to install the
special environment \emph{RProtoBuf:DescriptorPool} in the R search path.
The environment is special in that, instead of being associated with a
static hash table, it is dynamically queried by R as part of R's usual
variable lookup. In other words, it means that when the R interpreter
looks for a binding to a symbol (foo) in its search path,
it asks to our package if it knows the binding "foo", this is then
implemented by the \texttt{RProtoBuf} package by calling an internal
method of the \texttt{protobuf} C++ library.

\section{64-bit integer issues}
\label{sec:int64}

R does not have native 64-bit integer support.  Instead, R treats
large integers as doubles which have limited precision.  For example,
it loses the ability to distinguish some distinct integers:

<<>>=
2^53 == (2^53 + 1)
@

Protocol Buffers are frequently used to pass data between different
systems, however, and most other systems these days have support for
64-bit integers.  To work around this, RProtoBuf allows users to get
and set 64-bit integer types by treating them as characters.

<<echo=FALSE,print=FALSE>>=
if (!exists("protobuf_unittest.TestAllTypes",
            "RProtoBuf:DescriptorPool")) {
    unittest.proto.file <- system.file("unitTests", "data",
                                       "unittest.proto",
                                       package="RProtoBuf")
    readProtoFiles(file=unittest.proto.file)
}
@

If we try to set an int64 field in R to double values, we lose
precision:

<<>>=
test <- new(protobuf_unittest.TestAllTypes)
test$repeated_int64 <- c(2^53, 2^53+1)
length(unique(test$repeated_int64))
@

However, we can specify the values as character strings so that the
C++ library on which RProtoBuf is based can store a true 64-bit
integer representation of the data.

<<>>=
test$repeated_int64 <- c("9007199254740992", "9007199254740993")
@

When reading the value back into R, numeric types are returned by
default, but when the full precision is required a character value
will be returned if the \texttt{RProtoBuf.int64AsString} option is set
to \texttt{TRUE}.

<<>>=
options("RProtoBuf.int64AsString" = FALSE)
test$repeated_int64
length(unique(test$repeated_int64))
options("RProtoBuf.int64AsString" = TRUE)
test$repeated_int64
length(unique(test$repeated_int64))
@

<<echo=FALSE,print=FALSE>>=
options("RProtoBuf.int64AsString" = FALSE)
@

\section{Other approaches}

Saptarshi Guha wrote another package that deals with integration
of protocol buffer messages with R, taking a different angle :
serializing any R object as a message, based on a single catch-all
\texttt{proto} file.  Saptarshi's package is available at
\url{http://ml.stat.purdue.edu/rhipe/doc/html/ProtoBuffers.html}.

Jeroen Ooms took a similar approach influenced by Saptarshi in his
\texttt{RProtoBufUtils} package.  Unlike Saptarshi's package,
RProtoBufUtils depends on RProtoBuf for underlying message operations.
This package is available at
\url{https://github.com/jeroenooms/RProtoBufUtils}.

% Phillip Yelland wrote another implementation, currently proprietary,
% that has significant speed advantages when querying fields from a
% large number of protocol buffers, but is less user friendly for the
% basic cases documented here.

@@

Unlike Apache Thrift, Protocol Buffers do not include a concrete RPC
implementation.  However, serialized protocol buffers can trivially be
sent over TCP or integrated with a proprietary RPC system.  Combined
with an RPC system this means that one can interactively craft request
messages, send the serialized message to a remote server, read back a
response, and then parse the response protocol buffer interactively.

@@

Its pretty useful.  Murray to see if he can get approval to talk a
tiny bit about how much its used at Google.

%This file is only a basic article template. For full details of \emph{The R Journal} style and information on how to prepare your article for submission, see the \href{http://journal.r-project.org/latex/RJauthorguide.pdf}{Instructions for Authors}.
